Predicting the effects of character displacement on food-web dynamics

### 2020-07-02 manuscript_ECD_model.tex
incorporate reviewer and editor comments.

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{McCann2005}. The
second scenario accounts for this spatial context (Fig.
\ref{fig:foraging_scenarios}b) and takes the form:

@@

Given that character displacement occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when there is no
change in the population growth rates of consumers and resources (i.e.,
the rates of change in equation \ref{eq:1} are 0), and solving the
system at this point gives equilibrium abundances for each resource
(\(\hat R_i\)) and each consumer (\(\hat C_j\)). I also compared the
local stability of these food webs using standard methods
\citep{Otto2007}. This stability analysis derives the dominant
eigenvalue, \(\lambda_{\text{max}}\), of the matrix of partial
derivatives of each species' population growth rate (given by equation
\ref{eq:1}) with respect to each species' abundance evaluated at
equilibrium. If \(-\lambda_{max}>0\), then the food web will return to
equilibrium after a small perturbation (i.e., it is locally stable),
with more positive values indicating a faster return time. If
\(-\lambda_{max}<0\), then the food web is not locally stable.

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in Fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line in Fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in Fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.,
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub and has been archived with Zenodo (please contact
journal office for links).

@@

The effect of character displacement on resources corresponds to its
impact on food-web stability. For example, when character displacement
decreases resource abundances (orange points in Fig.
\ref{fig:plot_fig3}b), there is also a decrease in food-web stability
(Fig. \ref{fig:plot_fig3}c). Character displacement may not affect or
even increase food-web stability (blue and green lines in Fig.
\ref{fig:plot_fig3}c); however, evolution does not favor strong
divergence in these scenarios (blue and green points in Fig.
\ref{fig:plot_fig3}a), which dampens these contingent effects. Note that
the dip in stability occurs when both consumers evolve to be
generalists, a situation that is not favored in any of the foraging
scenarios I examined (Fig. \ref{fig:plot_fig3}c).

@@

As seen previously, the effect of character displacement on resource
abundances qualitatively corresponds to its effect on food-web stability
(Fig. \ref{fig:plot_fig3}f). Specifically, character displacement
decreases food-web stability, regardless of the tradeoff in attack
rates. This is not simply a consequence of having an additional consumer
in the system, but emerges from the eco-evolutionary feedback between
character displacement and resource suppression (Fig.
\ref{fig:plot_fig3}c). For example, when the tradeoff is concave up
(orange), the initial two-consumer food web (small circle) is more
stable than when there is only one consumer (small triangle); however,
this pattern switches by the end of the eco-evolutionary simulation
(large points).

\begin{figure}
\centering
\includegraphics{Fig_3_MacArthur_LawlorSmith.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1} (the choice to display
\emph{C}\textsubscript{1} was arbitrary), whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for the effective attack rate is different.}
\end{figure}

@@

In the first two foraging scenarios, character displacement influences
food-web stability, but all of the food webs ultimately return to a
stable equilibrium (because \(-\lambda_{max}>0\), see Appendix S2-3). In
this more realistic model, however, whether the food web is locally
stable depends on consumer and resource parameters. Specifically, I
found that the two-consumer food web will transition from having a
locally stable equilibrium to a limit cycle under the following
conditions (derived using Routh-Hurwitz criteria in Appendix S1):

@@

One of the criteria used to demonstrate ecological character
displacement is that ``sites of sympatry {[}two consumers{]} and
allopatry {[}one consumer{]} should not differ greatly in food
\(\textbf{[}\)resource abundances\(\textbf{]}\)'' \citep{Schluter1992}.
In contrast, my results indicate that character displacement causes
predictable differences in resource abundances. In fact, the ecological
and evolutionary scenarios that favored the largest character
displacement always decreased the relative abundance of resources. For
example, if mobile consumers compete for resources that occur in
different habitats, then character displacement always resulted in lower
resource abundances. Threespine stickleback, one of the classic examples
of character displacement, exemplify this foraging scenario
\citep{Schluter1992, Schluter2000}. Stickleback must move between the
pelagic and littoral zones of a lake when foraging for zooplankton and
benthic invertebrates, respectively. The theory developed here predicts
that resource abundances will be lower in lakes where competing
stickleback have undergone character displacement compared to lakes with
only a single species of stickleback. Interestingly, a disproportionate
number of the documented cases of character displacement involve
carnivores \citep{Schluter2000} that are larger, and likely more mobile,
than their resources \citep{McCann2005}, suggesting that many cases of
ecological character displacement may result in lower resource
availability.

Similarly, the evolutionary tradeoff that favored character displacement
decreased resource availability across all foraging scenarios. Although
data on the shape of the tradeoff in consumer foraging traits is scarce,
two classic examples of character displacement, Darwin's finches and
threespine stickleback, both appear to exhibit a tradeoff where extreme
trait values increase the net foraging rate of consumers
\citep{Schluter1985, Arnegard2014}. While it is theoretically possible
that character displacement does not alter (or even increase) resource
abundances, this was limited to the simplest, and arguably least
realistic, foraging scenario and under tradeoffs that did not favor
large displacements, and thus less likely to detect in nature. Taken
together, my results call for empirical work to test these clear
theoretical predictions and suggest a revision is needed for one of the
criteria used to demonstrate character displacement.

@@

A handful of empirical patterns support the hypothesis that character
displacement decreases food-web resilience. For example, a single
species of threespine stickleback lives in hundreds of small coastal
lakes in British Columbia, but the species pair, where character
displacement has resulted in specialized limnetic and benthic species,
are only known from six lakes in four independent watersheds
\citep{Schluter1992, Schluter2000}. Perhaps many lakes had a species
pair in the past, but have lost a species due to a less resilient
food-web structure \citep{Borrelli2015a, Borrelli2015b}. The species
pair are known to be vulnerable to perturbations, as they have gone
extinct in two of the six lakes after the introduction of nonnative
species \citep{Hatfield2001, Taylor2006, Rudman2016}. The vulnerability
of the stickleback system also corresponds with the fact that aquatic
food webs have several properties that make them less resilient to
perturbations, such as higher productivity and more efficient energy
transfer to consumers \citep{Rip2011}. Detecting the ghost of
competition past \citep{Connell1980} may be quite difficult, but it
could be possible with recent advances in genomics. For example,
\citet{Feulner2018} detected genomic signatures of hybridization in
sympatric whitefish species following periods of eutrophication. Perhaps
solitary stickleback in some lakes retain genomic signatures of having
been a habitat specialist in the past.

@@

@@

I studied this eco-evolutionary feedback between consumers and resources
using an Adaptive Dynamics approach. A strength of this approach is that
it enabled me to gain analytical insight to the effects of character
displacement in a more realistic foraging scenario. This is much less
tractable in quantitative genetic \citep{Taper1985, McPeek2017} or
explicit genetic \citep{Doebeli1996} models of character displacement,
which is why the foraging scenarios previously examined have been
limited \citep[but see][]{McPeek2017}. A weakness, however, is that I
assume a separation of time scales between ecological and evolutionary
dynamics, an assumption that is becoming less tenable
\citep{Hairston2005, Hendry2016}. I also do not explicitly model an
underlying phenotypic trait for consumer attack rates nor do I allow for
intraspecific variation. That being said, my theoretical predictions are
likely robust to these assumptions. This is because models that
explicitly include resource dynamics inevitably show that resource
competition results in character displacement, regardless of whether a
quantitative genetic or Adaptive Dynamics approach is used
\citep{Lawlor1976, Taper1985}. A quantitative genetic model may
certainly show differences in the pace of character displacement, but
this should not qualitatively change its effect on food-web dynamics. It
is important to note that my conclusions only apply to food webs with
biotic resources that are nutritionally substitutable. It would be
interesting to extend these current analyses to non-substitutable
resources where convergent character displacement is expected
\citep{Abrams1987, Fox2008}.

>>>>

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{Holt1984}. The second
scenario accounts for this spatial context (Fig.
\ref{fig:foraging_scenarios}b) and takes the form:

@@

Given that character displacement occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when there is no
change in the population growth rates of consumers and resources (i.e.,
the rates of change in equation \ref{eq:1} are 0), and solving the
system at this point gives equilibrium abundances for each resource
(\(\hat R_i\)) and each consumer (\(\hat C_j\)). I also compared the
local stability of these food webs using standard methods
\citep{Otto2007}. This stability analysis derives the dominant
eigenvalue, \(\lambda_{\text{max}}\), of the matrix of partial
derivatives of each species' population growth rate (given by equation
\ref{eq:1}) with respect to each species' abundance evaluated at
equilibrium. All eigenvalues must be negative for the food web to be
locally stable; however, the dominant eigenvalue,
\(\lambda_{\text{max}}\), is the one closest to positive infinity and
ultimately determines food-web stability. If \(\lambda_{\text{max}}>0\),
then the food web is not locally stable. If \(\lambda_{\text{max}}<0\),
then the food web will return to equilibrium after a small perturbation
(i.e., it is locally stable), with more negative values indicating a
faster return time (i.e., greater stability). I chose to plot
\(-\lambda_{\text{max}}\) rather than \(\lambda_{\text{max}}\), so that
increasing values would indicate greater stability, which I felt was a
more intuitive way to present the results.

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in Fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line in Fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in Fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.,
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0).

The Adaptive Dynamics approach I use here is one of several possible
approaches for studying the eco-evolutionary dynamics of interacting
species. A strength of this approach is that it enables me to gain
analytical insight to the effects of character displacement in more
realistic foraging scenarios. This is much less tractable in
quantitative genetic \citep{Taper1985, McPeek2017} or explicit genetic
\citep{Doebeli1996} models of character displacement, which is why the
foraging scenarios previously examined have been limited \citep[but
see][]{McPeek2017}. A weakness, however, is that I assume a separation
of time scales between ecological and evolutionary dynamics, an
assumption that is becoming less tenable
\citep{Hairston2005, Hendry2016}. In addition, I assume that there is no
intraspecific variation within interacting species, despite the fact
that intraspecific variation is a common feature of natural populations
and often has important ecological consequences
\citep{Bolnick2011, Des_Roches2018}. Still, models that explicitly
include resource dynamics inevitably show that resource competition
results in character displacement, regardless of whether a quantitative
genetic or Adaptive Dynamics approach is used
\citep{Lawlor1976, Taper1985}. Therefore, it should still serve as a
good starting point for studying the ecological consequences of
character displacement (but see \textbf{Caveats} section in the
discussion).

Unless otherwise noted, I conducted simulations with the following
parameter values: \emph{r} = 1; \emph{K} = 4; \emph{e} = 0.8; \emph{m} =
1; \emph{A} = 2; \emph{h} = 0.4; and \(w_{11} = w_{22} = 0.6\). I set an
initial value of \(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and
\(a_{21}\) depended on the value of \emph{n}. I set initial consumer and
resource abundances to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All
simulations were conducted in R \citep{R} and the code to reproduce
these simulations is publically available on GitHub and has been
archived with Zenodo (please contact journal office for links).

@@

The effect of character displacement on resources corresponds to its
impact on food-web stability. For example, when character displacement
decreases resource abundances (orange points in Fig.
\ref{fig:plot_fig3}b), there is also a decrease in food-web stability
(Fig. \ref{fig:plot_fig3}c). Character displacement may not affect or
even increase food-web stability (blue and green lines in Fig.
\ref{fig:plot_fig3}c); however, evolution does not favor strong
divergence in these scenarios (blue and green points in Fig.
\ref{fig:plot_fig3}a), which dampens these contingent effects. Note that
the dip in stability toward a dominant eigenvalue of zero is an artifact
of the two consumers becoming ecologically identical
(\(a_{11}=a_{21}=a_{12}=a_{22}\)) on identical resources, and
effectively become neutral species \citep{McPeek2019Neutral}. Although
neutrality is an ecological possibility, it is never favored by
evolution in these models since character displacement occurred across
all the foraging scenarios I examined.

@@

As seen previously, the effect of character displacement on resource
abundances qualitatively corresponds to its effect on food-web stability
(Fig. \ref{fig:plot_fig3}f). Specifically, character displacement
decreases food-web stability, regardless of the tradeoff in attack
rates. This is not simply a consequence of having an additional consumer
in the system, but emerges from the eco-evolutionary feedback between
character displacement and resource suppression (Fig.
\ref{fig:plot_fig3}c). For example, when the tradeoff is concave up
(orange), the initial two-consumer food web (small circle) is more
stable than when there is only one consumer (small triangle); however,
this pattern switches by the end of the eco-evolutionary simulation
(large points). As with the previous model, the dip in stability
corresponds to the situation where consumers are effectively identical
(\(w_{11}a_{11}=w_{21}a_{21}=w_{12}a_{12}=w_{22}a_{22}\)) and become
neutral species, which is never favored in these models.

It is worth noting that the scales are different between the left and
right columns of Figure \ref{fig:plot_fig3}. When resources occur in
different habitats (relative to the same habitat), the consumer's
effective attack rate is lower because it is multiplied by the
proportion \(w_{ij}\). This results in relatively higher resource
abundances and more stable food webs compared to when resources occur in
the same habitat, and corresponds to the inherent stabilizing role of
spatial heterogeneity in food webs \citep{Holt1984, McCann2005}.

\begin{figure}
\centering
\includegraphics{Fig_3_MacArthur_LawlorSmith.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1} (the choice to display
\emph{C}\textsubscript{1} was arbitrary), whereas small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for the effective attack rate is different.}
\end{figure}

@@

In the first two foraging scenarios, character displacement influences
food-web stability, but all of the food webs ultimately return to a
stable equilibrium (because \(-\lambda_{max}>0\), see Appendix S2-3). In
this more realistic model, however, whether the food web is locally
stable depends on consumer and resource parameters. Specifically, I
found that the two-consumer food web will become unstable under the
following conditions (derived using Routh-Hurwitz criteria in Appendix
S1):

@@

One of the criteria used to demonstrate ecological character
displacement is that ``sites of sympatry {[}two consumers{]} and
allopatry {[}one consumer{]} should not differ greatly in food
\(\textbf{[}\)resource abundances\(\textbf{]}\), climate, or other
environmental features affecting the phenotype'' \citep{Schluter1992}.
In contrast, my results indicate that character displacement causes
predictable differences in resource abundances. In fact, the ecological
and evolutionary scenarios that favored the largest character
displacement always decreased the relative abundance of resources. For
example, if mobile consumers compete for resources that occur in
different habitats, then character displacement always resulted in lower
resource abundances. Threespine stickleback, one of the classic examples
of character displacement, exemplify this foraging scenario
\citep{Schluter1992, Schluter2000}. Stickleback must move between the
pelagic and littoral zones of a lake when foraging for zooplankton and
benthic invertebrates, respectively. The theory developed here predicts
that resource abundances will be lower in lakes where competing
stickleback have undergone character displacement compared to lakes with
only a single species of stickleback. Interestingly, a disproportionate
number of the documented cases of character displacement involve
carnivores \citep{Schluter2000} that are larger, and likely more mobile,
than their resources \citep{McCann2005}, suggesting that many cases of
ecological character displacement may result in lower resource
availability.

Similarly, the evolutionary tradeoff that favored character displacement
decreased resource availability across all foraging scenarios. Although
data on the shape of the tradeoff in consumer foraging traits is scarce,
two classic examples of character displacement, Darwin's finches and
threespine stickleback, both appear to exhibit a tradeoff where extreme
trait values increase the net foraging rate of consumers
\citep{Schluter1985, Arnegard2014}. While it is theoretically possible
that character displacement does not alter (or even increase) resource
abundances, this was limited to the simplest, and arguably least
realistic, foraging scenario and under tradeoffs that did not favor
large displacements, and thus less likely to detect in nature.

My predictions could be tested with an experiment that puts consumers
that have and have not undergone character displacement in a common
resource setting and comparing their effect on resource abundances. One
experiment with threespine stickleback has actually done this, although
they focused more on the community composition of zooplankton and
benthic invertebrates as well as ecosystem properties
\citep{Harmon2009}. Their reported data on average resource abundances
support my prediction that character displacement in stickleback would
decrease total resource abundances \citep[compare totals for ``BL'' vs.
``G'' in Table 1 of][]{Harmon2009}. For future tests, I want to
emphasize that it is important to measure resource abundances over time,
as I also predict that abundances will be more variable (see
\textbf{Food-web stability} section). I specifically predict that
character displacement would decrease time-averaged resource abundances.
An even more explicit test would be to apply a first-order multivariate
autoregressive model to obtain equilibrium estimates of resource
abundances \citep[see equation 15 in][]{Ives2003}. An advantage of using
this type of statistical model is that one can also estimate the
effective attack rate of consumers, which I predict to be increased by
character displacement in most foraging scenarios. In designing an
experiment to test this theory, it is important to keep in mind the
amount of available space for consumers. For example, a mesocosm
experiment may effectively constrain resources to be in the same habitat
for the consumer, in which case I predict that the shape of the tradeoff
in consumer attack rates will determine the impact of character
displacement (see Fig. \ref{fig:plot_fig3}a-c).

Although a well-designed experiment would be an ideal test, the same
methods described above could be applied to observational data from the
field. In fact, field data may already be available that could be used
to test the effects of character displacement \citep{Schluter2000}. For
example, the criterion that my theory questions is the third most
frequently assessed of the six criteria \citep[see ``Environmental
controls'' in Figure 1 of][]{Schluter2000}. The data from studies that
measured resources could be used to test my prediction that character
displacement decreases resource abundances. Although my theory calls
into question part of this criterion, I agree with its intention, which
was to ensure that character displacement was not due to species
evolving independently to different environments \citep{Schluter1992}.

@@

@@

Testing the effect of character displacement on food-web stability will
be inherently more difficult than for resource abundances. Fortunately,
a common empirical metric of stability, the coefficient of variation in
abundance (CV), corresponds to changes in the dominant eigenvalue of
consumer-resource interactions \citep{Gellner2016TheorEcol}. The
coefficient of variation is an indicator of temporal variability and
requires data on the mean (\(\mu\)) and standard deviation (\(\sigma\))
of population abundance over time (\(\text{CV}=\sigma/\mu\)). I predict
that character displacement would increase the coefficient of variation
(across most foraging scenarios) in each consumer and resource
population. As with resource abundances, a more explicit test would be
to apply multivariate autoregressive models to time series of consumer
and resource abundances in order to directly estimate the dominant
eigenvalue of the food web \citep[see equation 22 in][]{Ives2003}.

@@

One assumption that warrants further consideration is that I do not
allow for intraspecific variation. Models of character displacement that
allow trait variation to evolve actually predict that consumers will
exhibit more trait variation in the absence of a competitor
\citep{Taper1985}. On the one hand, increased trait variation would
impose a greater fitness load for the single consumer, which would
weaken its top-down effect on resources \citep{Schreiber2011}. This
would act to magnify my predicted effects of character displacement on
resource abundances and food-web stability. On the other hand, if the
increased trait variance has a large heritable component, then rapid
evolution may generate oscillatory or chaotic dynamics in the single
consumer food web \citep{Schreiber2011}, which could negate my predicted
effects of character displacement on food-web stability. Finally, it is
important to note that my conclusions only apply to food webs with
biotic resources that are nutritionally substitutable. It would be
interesting to extend these current analyses to non-substitutable
resources where convergent character displacement is expected
\citep{Abrams1987, Fox2008}.
